On Misuses of the Kolmogorov--Smirnov Test for One Sample Goodness-of-Fit

### 2022-04-28 Manuscript/manuscript.tex
small fix
various edits

\usepackage[pagewise]{lineno}
\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

\title{On Misuses of the Kolmogorov–-Smirnov Test for One Sample Goodness-of-Fit}

\author{Anthony Zeimbekakis and Jun Yan\\
\href{mailto:anthony.zeimbekakis@uconn.edu}{\nolinkurl{anthony.zeimbekakis@uconn.edu}}\\
Department of Statistics, University of Connecticut}
\date{April 21, 2022}

@@

The Kolmogorov-Smirnov (KS) statistic is one of the most popular goodness-of-fit 
tests for comparing a sample with a hypothesized parametric distribution.
Let $X_1, ..., X_n$ be a random sample from some continuous distribution. 
The null hypothesis $H_0$ is that the population distribution is $F$.
Let $F_n(\cdot)$ be the empirical cumulative distribution. The KS statistic is
\[
  D = \sup_x {\lvert F_{n}(x) - F(x) \rvert}.
\]
The distribution of $D$ under $H_0$ turned out to be independent of the
distribution $F$ and a table of critical values for $D$ has been constructed
\citet{Massey} for various sample sizes $n$ and significance 
levels $\alpha$. If the value of $d$ exceeds the test's corresponding critical 
value, the null hypothesis is rejected. The KS test is available in popular 
statistical software packages, such as function \texttt{ks.test} in R 
\citep{R, Marsaglia}.

@@

The KS test is not applicable when the hypothesized distribution contains fitted
parameters. \citet{Steinskog} demonstrates the change in power when using 
estimated parameters and stress caution in using the KS test in such ways. 
\citet{Lilliefors} shows that using the standard table when values of the mean 
and standard deviation are estimated obtains extremely conservative results. 
This is supported by \citet{Capasso}, which concludes that failing to 
re-estimate the parameters may lead to wrong, overly-conservative approximations 
to the distributions of goodness-of-fit test statistics based on the empirical 
distribution function. \citet{Capasso} also notes that the impact of this 
possible mistake may turn out to be dramatic and does not vanish as the sample 
size increases. Remedies are provided by \citet{Genest} and \citet{Babu} in the 
form of bootstrap. \citet{Genest} provides validity for using parametric 
bootstrap with various goodness-of-fit tests. \citet{Babu} details the bootstrap
procedure for goodness-of-fit tests and notes that both parametric and 
non-parametric procedures lead to correct asymptotic levels, however there 
is a correction required for the non-parametric case.

@@

The contribution of this paper is a demonstration of misuses of the one-sample
KS test in three scenarios and their remedies in practice. In order to set 
up the demonstrations, simulated data is used throughout. The random data is 
generated with sample size $n = 100$ and the remedies are performed on various
families of distributions.

@@

The null distribution of the KS statistic changes when the hypothesized
distribution contains fitted parameters. In this scenario, the null hypothesis
is $H_0$: the random sample $X_1, \ldots, X_n$ come from a continuous
distribution $F_{\theta}$ with unspecified parameter $\theta$.
Let $\hat\theta_n$ be an estimator of $\theta$, which could be, for example,
maximum likelihood estimator of moment estimator. The test statistic is
\[
  D = \sup_x | F_n(x) - F_{\hat\theta_n}(x) |.
\]
Because $F_{\hat\theta_n}$ is not the same as the true data generating
$F_\theta$, the null distribution of $D$ obtained in existing implementations in
software packages which assumes completely known $F_\theta$ no long applies.
\jy{cite the paper that ``rediscovered'' the fact}

@@

To fix the problem, parametric bootstrap can be used to approximate the null
distribution of the testing statistic. 
\begin{enumerate}
  \item 
    Draw a random sample $X_1^*,...,X_n^*$ from the fitted distribution $F_{\hat\theta_n}$
  \item 
    Fit $F_\theta$ to the sample and obtain estimated $\hat\theta_n^*$
  \item
    Obtain the empirical distribution function of the random sample $F_n^*$
  \item 
    Calculate bootstrap KS statistic
    \[
      D^* = \sup_x \lvert F_n^* (x)- F_{\hat\theta_n}^*(x) \rvert.
    \]
  \item
    Repeat the previous steps a large number $B$ times and use the empirical
    distribution of the $B$ test statistics as an approximate distribution of
    the null distribution of the observed statistic.      
\end{enumerate}

% In parametric bootstrap, the resampling is done where $\hat{F}_n = F(.;\hat{\theta}_n)$. 
% Bootstrap samples are drawn from ${F}(.,\hat{\theta}_n)$ where the assumed distribution 
% $F$ is normal and $\hat{\theta}_n = (\bar X, s)$, the sample mean and sample standard 
% deviation. For each bootstrap sample, a KS test statistic is found using those 
% fitted parameters.

\jy{for clarity, give the steps of the nonparametric bootstrap procedure and
  make clear the bias correction step.}

In non-parametric bootstrap, the resampling is done where $\hat{F}_n = F_n$. The non-parametric
case requires a correction for bias detailed by \citet{Babu}. There is a known bias
term $B_{n}(x) = \sqrt{n}(F_{n}(x) - F(x;\hat{\theta}_n)$. Including this term in the
calculation of the KS statistic corrects the bias \citet{Babu}:
$d = \max\lvert F_{n}(x) - F(x) - B_{n}(x) \rvert$.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{hist_fitted}
  \caption{Histograms demonstrating fitted parameters. \jy{make the caption self-contained.}}
  \label{fig:hist_fitted}
\end{figure}

Figure~\ref{fig:hist_fitted} displays the results of from our simulations. The procedure 
is replicated $1000$ times using the standard normal distribution with sample size $n=100$. 
In each replicate test, $1000$ bootstrap samples are obtained. The p-value can be calculated
by counting the number of bootstrap KS statistics greater than or equal to the observed KS statistic, 
and then dividing by the number of bootstrap samples. It is clear from the figure that both
bootstrap processes correct for the problem of fitted parameters. The plots appear to be
$U(0,1)$, which is expected of a full power test.

@@

The KS test also displays issues in the case of dependence. As mentioned, an assumption of the 
test is that the data is independent. Unfortunately, real data is often temporally
or spatially dependent and the results of a goodness-of-fit test would be valuable. 
When the KS test is performed on dependent data it performs poorly. This is demonstrated
with a simulation in Figure~\ref{fig:hist_correlation}. Data is generated from a
\jy{give more details about the simulation setting: what ar1 coefficients were
  considered? What was the null hypothesis? How was the test statistic calculated?}
first-order autoregressive model (AR(1)) with a standard normal distribution. The simulation
is done with the levels of $\psi$ varying from $(-3,3)$. The results echo those of \citet{Durilleul}
that the KS statistic is too liberal for positive autocorrelation values, and 
that the behavior is asymmetrical for negative values.

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_correlation}
  \caption{Histograms of P-values with correlated data.}
  \label{fig:hist_correlation}
\end{figure}

In order to correct this, we can employ a parametric bootstrap procedure which
assumes a working serial dependence structure through copulas. A copula is a
multivariate distribution with standard uniform marginal distributions, which
completely characterizes the dependence structure of a multivariate
distribution \cite{copula, book}. For simplicity, we assume a normal copula with
an AR-1 structure to characterize the serial dependence of the observations. The
AR-1 parameter of the normal copula is set to match the be the sample serial
Spearman's rho of the observed sample. The procedure is as follows.

\begin{enumerate}
  \item Calculate the observed KS statistic \jy{the observed statistic is
      calculated the same way still, so it is not needed in the algorithm here}
  \item Get the lag-1 sample auto-spearman rho
  \item Perform parametric bootstrap to get an empirical distribution of the KS
    statistic
    \jy{need to spell out the details; how to generate $X_t$ given $X_{t-1}$?}
\end{enumerate}

@@

Figure~\ref{fig:hist_copula_only} displays the results of the copula remedy for dependent data.
The data is generated from an AR(1) model with $\psi = .5$. The copula used is to model
dependence is the normal copula. $1000$ replicate tests were performed with
$1000$ bootstrap samples per test.
\jy{present other scenarios too}
The naive test is performed by simply using the KS test without correcting for any 
potential dependence.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_ma1}
  \caption{Histograms demonstrating remedy use with MA(1) model.}
  \label{fig:hist_copula_ma1}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_arma}
  \caption{Histograms demonstrating remedy use with ARMA(1,1) model.}
  \label{fig:hist_copula_arma}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_gamma}
  \caption{Histograms demonstrating remedy with Gamma(3,1).}
  \label{fig:hist_copula_gamma}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_gev}
  \caption{Histograms demonstrating copula use to correct dependence.}
  \label{fig:hist_copula_gev}
\end{figure}

Where this remedy falls short is when the approximation of an AR(1) model is too inaccurate.
An AR(2) model can be used to demonstrate this. For an AR(2) model with $\phi = (0.5, 0.3)$,
Figure~\ref{fig:hist_copula_ar2} displays the results of applying the various remedies. 
While the remedies improve the plot of the p-values in comparison to the naive case,
because the approximation is not close enough the correction does not work completely.

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_ar2}
  \caption{Histograms demonstrating copula use to correct dependence.}
  \label{fig:hist_copula_ar2}
\end{figure}

@@

The KS test has base assumptions that the hypothesized distribution is completely specified
and the data is independent. When these assumptions are violated, the test is no
longer accurate and remedies must be performed. In the case of fitted parameters, 
parametric and non-parametric bootstrap can restore the power of the test. A bias 
correction is required if the non-parametric form is used \citep{Babu}. In the case of dependent
data, a procedure using bootstrap with copulas to model dependency shows positive results.
When both assumptions are violated, that is the case where the data has dependence and 
parameters must be fitted, the copula remedy also shows positive results. The tests were 
performed on simulated data from a standard normal distribution, though also appear effective
against other distributions such as gamma and gev. The copula remedy
is not a complete solution. Regardless of the true dependence, we use an AR(1) model.
Therefore, if the AR(1) model is a close approximation of the truth, the fix can work.
However, in cases such as AR(2), the approximation is incorrect and the fix does not 
completely restore power.

>>>>

\usepackage[pagewise]{lineno}
%\linenumbers*[1]
% %% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
 {\linenomath\csname old#1\endcsname}%
 {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%

@@

\title{On Misuses of the Kolmogorov--Smirnov Test for One Sample Goodness-of-Fit}

\author{Anthony Zeimbekakis\\
\href{mailto:anthony.zeimbekakis@uconn.edu}{\nolinkurl{anthony.zeimbekakis@uconn.edu}}\\
Department of Statistics, University of Connecticut\\\\
Thesis Advisor: Jun Yan\\
Honors Advisor: Elizabeth Schifano}
\date{}

@@

The Kolmogorov-Smirnov (KS) statistic is one of the most popular goodness-of-fit 
tests for comparing a sample with a hypothesized parametric distribution.
Let $X_1, ..., X_n$ be a random sample from some continuous distribution. 
The null hypothesis $H_0$ is that the population distribution is $F$.
Let $F_n(\cdot)$ be the empirical cumulative distribution. The KS statistic is
\[
  D = \sup_x | F_{n}(x) - F(x) |.
\]
The distribution of $D$ under $H_0$ turned out to be independent of the
distribution $F$ and a table of critical values for $D$ has been constructed
\citep{Massey} for various sample sizes $n$ and significance 
levels $\alpha$. If the value of $d$ exceeds the test's corresponding critical 
value, the null hypothesis is rejected. The KS test is available in popular 
statistical software packages, such as function \texttt{ks.test} in R 
\citep{R, Marsaglia}.

@@

The standard KS test is not applicable when the hypothesized distribution 
contains fitted parameters. \citet{Steinskog} demonstrates the change in power 
when using estimated parameters and stress caution in using the KS test in such 
ways. \citet{Lilliefors} shows that using the standard table when values of the 
mean and standard deviation are estimated obtains extremely conservative 
results. This is supported by \citet{Capasso}, which concludes that failing to 
re-estimate the parameters may lead to wrong, overly-conservative approximations 
to the distributions of goodness-of-fit test statistics based on the empirical 
distribution function. \citet{Capasso} also notes that the impact of this 
possible mistake may turn out to be dramatic and does not vanish as the sample 
size increases. Remedies are provided by \citet{Genest} and \citet{Babu} in the 
form of bootstrap. \citet{Genest} provides validity for using parametric 
bootstrap with various goodness-of-fit tests. \citet{Babu} details the bootstrap
procedure for goodness-of-fit tests and notes that both parametric and 
non-parametric procedures lead to correct asymptotic levels, however there 
is a correction required for the non-parametric case.

@@

The contribution of this paper is a demonstration of misuses of the one-sample
KS test in three scenarios and their remedies in practice. The scenarios are 
where:
1) the hypothesized distribution has unspecified parameters;
2) the data are serially dependent; and
3) a combination of the first two scenarios. 
In each scenario, the misuse is performed and the impacts are shown. Then, a
remedy is detailed and performed alongside the misuse to show its positive 
effects. In order to set up the demonstrations, simulated data is used 
throughout. The remedies are also performed on a various families of 
distributions.

@@

The null distribution of the KS statistic changes when the hypothesized
distribution contains fitted parameters. In this scenario, the null hypothesis
is $H_0$: the random sample $X_1, \ldots, X_n$ come from a continuous
distribution $F_{\theta}$ with unspecified parameter $\theta$.
Let $\hat\theta_n$ be an estimator of $\theta$, which could be, for example,
maximum likelihood estimator of moment estimator. The test statistic is
\[
  D = \sup_x | F_n(x) - F_{\hat\theta_n}(x) |.
\]
Since $F_{\hat\theta_n}$ is not the same as the true data generating
$F_\theta$, the null distribution of $D$ obtained in existing implementations in
software packages which assumes completely known $F_\theta$ no long applies
\citep{Steinskog}.

@@

To fix the problem, parametric bootstrap can be used to approximate the null
distribution of the testing statistic. 
\begin{enumerate}
  \item 
    Draw a random sample $X_1^*,...,X_n^*$ from the fitted distribution $F_{\hat\theta_n}$
  \item 
    Fit $F_\theta$ to the sample and obtain estimated $\hat\theta_n^*$
  \item
    Obtain the empirical distribution function of the random sample $F_n^*$
  \item 
    Calculate bootstrap KS statistic
    \[
      D^* = \sup_x | F_n^* (x)- F_{\hat\theta_n}^*(x) |.
    \]
  \item
    Repeat the previous steps a large number $B$ times and use the empirical 
    distribution of $D^*$ to approximate the null distribution of the observed 
    statistic.    
\end{enumerate}

Nonparametric bootstrap can also be used to approximate the null distribution 
of the testing statistic. The procedure is similar, however there is a 
correction for bias that is required \citep{Babu}.
\begin{enumerate}
  \item 
    Draw a random sample $X_1^*,...,X_n^*$ from the empirical distribution $F_n$
    with replacement
  \item 
    Fit $F_\theta$ to the sample and obtain estimated $\hat\theta_n^*$
  \item
    Obtain the empirical distribution function of the random sample $F_n^*$
  \item 
    Calculate bootstrap KS statistic
    \[
      D^* = \sup_x | F_n^* (x)- F_{\hat\theta_n}^*(x) - B_n(x) |.
    \]
    where $B_{n}(x) = \sqrt{n}(F_{n}(x) - F_{\hat\theta_n}(x))$ is the known bias
    term \citep{Babu}
  \item
    Repeat the previous steps a large number $B$ times and use the empirical 
    distribution of $D^*$ to approximate the null distribution of the observed 
    statistic.
\end{enumerate}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{hist_fitted}
  \caption{Implementing bootstrap remedies for fitted parameters misuse. \jy{make the caption self-contained.}}
  \label{fig:hist_fitted}
\end{figure}

Figure~\ref{fig:hist_fitted} displays the results of from our simulations. The 
procedure is replicated $1000$ times using the standard normal distribution with
sample size $n=100$. In each replicate test, $B = 1000$ bootstrap samples are 
obtained. The p-value can be calculated by counting the number of bootstrap KS 
statistics greater than or equal to the observed KS statistic, and then dividing 
by the number of bootstrap samples. It is clear from the figure that both
bootstrap processes correct for the problem of fitted parameters. The plots 
appear to be $U(0,1)$, which is expected of a full power test.

@@

The KS test also displays issues in the case of dependence. As mentioned, an 
assumption of the test is that the data is independent. Unfortunately, real data
is often temporally or spatially dependent and the results of a goodness-of-fit 
test would be valuable. When the KS test is performed on dependent data it 
performs poorly. In this situation, the null hypothesis is $H_0$: the random 
sample $X_1, \ldots, X_n$ come from a continuous distribution $F_{\theta}$ with 
specified parameter $\theta$.
Let $\hat\theta_n$ be an estimator of $\theta$, which could be, for example,
maximum likelihood estimator of moment estimator. The test statistic is
\[
  D = \sup_x | F_n(x) - F_{\hat\theta_n}(x) |.
\]
However, since the data is dependent, the null hypothesis is wrong and must be 
corrected. This is demonstrated with a simulation in 
Figure~\ref{fig:hist_correlation}. Data is generated from a first-order 
autoregressive model (AR-1) with a standard normal distribution. The simulation
is done with the levels of the AR coefficient $\psi$, varying from $(-3,3)$. 
The results echo those of \citet{Durilleul} that the KS statistic is too liberal 
for positive autocorrelation values, and that the behavior is asymmetrical for 
negative values.

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_correlation}
  \caption{Results of varying AR levels of AR-1 simulated data.}
  \label{fig:hist_correlation}
\end{figure}

In order to correct this, we can employ a parametric bootstrap procedure which
assumes a working serial dependence structure through copulas. A copula is a
multivariate distribution with standard uniform marginal distributions, which
completely characterizes the dependence structure of a multivariate
distribution \citep{Copula, Hofert}. For simplicity, we assume a normal copula 
with an AR-1 structure to characterize the serial dependence of the observations. 
The AR-1 parameter of the normal copula is set to match the be the sample serial
Spearman's rho of the observed sample. The procedure is as follows.

\begin{enumerate}
  \item 
    Black box
    \jy{need to spell out the details; how to generate $X_t$ given $X_{t-1}$?}
  \item 
    Transform bootstrap sample to 
  \item
    Repeat the previous steps a large number $B$ times and use the empirical
    distribution of the $B$ test statistics as an approximate distribution of
    the null distribution of the observed statistic. 
\end{enumerate}

@@

Figure~\ref{fig:hist_copula_only} displays the results of the copula remedy for dependent data.
The data is generated from an AR-1 model with $\psi = .5$. The copula used is to model
dependence is the normal copula. $1000$ replicate tests were performed with
$B = 1000$ bootstrap samples per test.
\jy{present other scenarios too}
The naive test is performed by simply using the KS test without correcting for any 
potential dependence.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_ma1}
  \includegraphics{hist_copula_arma}
  \caption{MA-1 and ARMA(1,1)}
  \label{fig:hist_copula_ma1}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_gamma}
  \includegraphics{hist_copula_gev}
  \caption{Gamma(3,1)}
  \label{fig:hist_copula_gamma}
\end{figure}

Where this remedy falls short is when the approximation of an AR-1 dependence 
structure is too inaccurate. An AR-2 model can be used to demonstrate this. For 
an AR-2 model with $\phi = (0.5, 0.3)$, Figure~\ref{fig:hist_copula_ar2} 
displays the results of applying the various remedies. While the remedies 
improve the plot of the p-values in comparison to the naive case, because the 
approximation is not close enough the correction does not work completely.

\begin{figure}[tbp]
  \centering
  \includegraphics{hist_copula_ar2}
  \caption{AR-2}
  \label{fig:hist_copula_ar2}
\end{figure}

@@

The KS test has base assumptions that the hypothesized distribution is 
completely specified and the data is independent. When these assumptions are 
violated, the test is no longer accurate and remedies must be performed. In the 
case of fitted parameters, parametric and non-parametric bootstrap can restore 
the power of the test. A bias correction is required if the non-parametric form 
is used \citep{Babu}. In the case of dependent data, a procedure using bootstrap 
with copulas to model dependency shows positive results. When both assumptions 
are violated, that is the case where the data has dependence and parameters must 
be fitted, the copula remedy also shows positive results. The tests were 
performed on simulated data from a standard normal distribution, though also 
appear effective against other distributions such as gamma and gev. The copula 
remedy is not a complete solution. Regardless of the true dependence, we use an 
AR-1 model. Therefore, if the AR-1 model is a close approximation of the truth, 
the fix can work. However, in cases such as AR-2, the approximation too far and 
the fix does not completely restore power.
